Classless Java: Tuning Java Interfaces

### 2015-11-24 paper/sections/RelatedWork.tex
related work ready for review (except for S8.4)

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  It provides
expressive power, but it is difficult to model and implement, and can create
programs that are hard to reason about.  These difficulties include the famous
diamond (fork-join) problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.
%and the yo-yo problem~\cite{taenzer1989problems}.  
To conciliate the need for
expressive power and the need for simplicity, lots of models have been proposed
in the past few years, including mixins~\cite{bracha90mixin} and
traits~\cite{scharli03traits}.  They provide novel programming architecture
models in the OO paradigm.

\begin{enumerate}
\item Mixins allows to name components that can be applied to various classes as
  reusable functionality units. However, they suffer from linearisation: the
  order of mixin application is relevant in often subtitle and undesired
  ways. constraints. This hinders their usability and their ability of resolving
  conflicts: the linearisation (total ordering) of mixin inheritance cannot
  provide a satisfactory resolution in all cases and restricts the flexibility
  of mixin composition.  To fight those limitations, an algebra of mixin
  operators is introduced~\cite{ancona2002calculus}, but this raised the
  complexity of the approach, especially when constructors and fields are
  considered~\cite{marco09FJigsaw}.  Our approach does not have the
  linearisation problem, in that the semantics of Java \textbf{extends} clause
  is unordered and symmetric.
\item Simplifying the mixin algebras approach, traits draw a strong line between
  units of reuse (traits) and object factories (classes) In this model,
  traits~\cite{scharli03traits} units of reusable code, containing only methods
  as reusable functionalities. Thus, no state/state initialization is
  considered.

  Classes act as object factories, requiring functionalities from multiple
  traits.  Traits offers a trait algebra with operations like sum, alias and
  exclusion, provided for explicit conflict resolution.

  Concluding, (pure) traits do not allow state and they do not offers any reuse
  instrument to ensure that state is coherently initialized when finally defined
  in classes.  Traits can't be instantiated. This model requires two concepts
  (traits and classes) to coexist and cooperate.

  Some authors see this as good language design fostering good software
  development by helping programmers to think about the structure of their
  programs.  However, other authors see the need of two concepts and the absence
  of state as drawbacks of this model. Our approach takes interfaces as units of
  reuse, and meanwhile generates factory method for instantiation.

\item C++ and Scala also try to provide solutions to multiple inheritance, but
  both suffer from object initialization problems. Virtual inheritance in C++
  provides another solution to multiple inheritance (especially the diamond
  problem by keeping only one copy of the base class)~\cite{ellis1990annotated},
  however suffers from object initialization problem as pointed out by Malayeri
  et al.~\cite{malayeri2009cz}. It bypasses all constructor calls to virtual
  superclasses, which would potentially cause serious semantic error. Scala
  solution (very similar to linearised mixins, but misleadingly called traits in
  the language) avoids this problem by disallowing constructor parameters,
  causing no ambiguity in cases such as diamond problem.  This approach has
  limited expressiveness, and suffers from all the problems of linearised mixin
  composition.
  Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
  \Q@ class A use B,C {...new methods...}@\\*
  can be expressed in python as \\*
  \Q@ class Aux: ...new methods...@\\*
  \Q@ class A(B,C,Aux): pass@
  Our approach not only does not involve linearised mixin problem, but also
  support proper constructor mechanism.
\end{enumerate}

@@

We can model certain types of language extensions with annotations 
only, but those extensions do not introduce new syntax: they 
merely do automatic code generation. 
\end{comment}
\subsection{Meta-programming competes with Language extensions}
The most obvious solution to add features to a language is language extension.
They are often implemented as syntactic extensions that can be desugared to the base
language. For example, the Scala compiler was extended to directly supports XML syntax. However, this approach does not support combining multiple extensions into one. We are de facto creating a fork in the language, and rarely the new fork gain enough traction to become the main language release.
On this topic we mention SugarJ~\cite{erdweg11sugarj}; a Java-based extensible language allowing programmers to extend it with custom features by definitions in meta-DSLs (SDF, Stratego, etc).

On the other side, when the starting language have a flexible enough syntax and a fast and powerful enough reflection, we may just need to play with operator overloading and other language tricks to discover that the language feature we need can be expressed as a simple library in our language. For example, consider SQL alchemy in python.

Java-like language tends to sits in the middle of this two extremes:
libraries can not influence the type system, so many solutions valid in python or other languages could not be applicable, or may be applicable at the cost of loosing safety.

Here (compile/load time) code generation come at the rescue: 
if for a certain feature (\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature
(annotating a class and provide getters), then we can insert in the compilation process a tool that exam and enrich the code before compilation. No need to modify the original source; for example we can work on temporary files.
Java is a particular good candidate for this kind of manipulation since it already provide ways to define and integrate such tools in its own compilation process: in this way there is no need of temporary files, and there is a well defined way of putting multiple extensions together.

Other languages offers even stronger support to safe code manipulation:
Template Haskell~\cite{}, F\# (type providers)~\cite{} and MetaFjig (Active Libraries)~\cite{}
all allows to execute code at compile time and to generate code/classes that are transparently integrated in the program that is being generated/processed/compiled.
In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or non nonsensical) would never generate ill-typed code. This is roughly equivalent to what we state and manually proof in Lemma 2 for our particular transformation.
Since MetaFjig is not working on annotated classes, there is no direct equivalence on the overall theorem of safety we shown.

\subsection{Formalization of Java8}
We provide a simple and well designed formalization for a subset of Java including default/static interface methods and object initialization literals (often called anonymous local inner classes).
A similar formalization was drafted by\marco{ I use the term draft because I seams to remember it was just a technical report, I'm right?}
Goetz and Field~\cite{goetz12fdefenders} to formalize defender (default) methods
in Java. However this formalization is limited to model exactly one
method inside classes/interfaces.

As a evidence of the attention and care present in our formalization work, while double-checking the behaviour of Java in side cases we have discovered a likely bug in the current \texttt{javac}.\marco{refer to before when we explain the issue.}

\section{Future work}

>>>>

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  It provides
expressive power, but it is difficult to model and implement, and can create
programs that are hard to reason about. These difficulties include the famous
diamond (fork-join) problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.
%and the yo-yo problem~\cite{taenzer1989problems}.  
To conciliate the need for expressive power and the need for simplicity, lots of
models have been proposed in the past few years, including C++ virtual
inheritance, mixins~\cite{bracha90mixin} and traits~\cite{scharli03traits}.
They provide novel programming architecture models in the OO paradigm. In terms
of restrictions set on these models, C++ virtual inheritance aims at a more
general model, mixins added some restrictions on the model, and trait model is
the most restricted one (excluding states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance. Virtual
inheritance in C++ provides another solution to multiple inheritance (especially
the diamond problem by keeping only one copy of the base
class)~\cite{ellis1990annotated}, however suffers from object initialization
problem as pointed out by Malayeri et al.~\cite{malayeri2009cz}. It bypasses all
constructor calls to virtual superclasses, which would potentially cause serious
semantic error.

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. It allows to name
components that can be applied to various classes as reusable functionality
units. However, they suffer from linearisation: the order of mixin application
is relevant in often subtitle and undesired ways. This hinders their usability
and the ability of resolving conflicts: the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in all cases and
restricts the flexibility of mixin composition. To fight those limitations, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raised the complexity of the approach, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala solution adopts linearised mixins,
(but misleadingly called traits in the language) avoids object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition.
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@

\noindent Our approach does not involve linearised mixin problem, in that the semantics of
Java \textbf{extends} clause is unordered and symmetric. Also,our approach
supports proper constructor mechanism.

\paragraph{Traits.}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse (traits) and object factories (classes). In this
model, traits as units of reusable code, contain only methods as reusable
functionalities. Thus, no state/state initialization is considered. Classes act
as object factories, requiring functionalities from multiple traits.  Traits
offer a trait algebra with operations like sum, alias and exclusion, provided
for explicit conflict resolution.  Concluding, (pure) traits do not allow state
and they do not offers any reuse instrument to ensure that state is coherently
initialized when finally defined in classes.  Traits can not be
instantiated. This model requires two concepts (traits and classes) to coexist
and cooperate.  Some authors see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model. Our approach takes interfaces as
units of reuse, and meanwhile generates factory method for instantiation.

% \begin{enumerate}
% \item Mixins allows to name components that can be applied to various classes as
%   reusable functionality units. However, they suffer from linearisation: the
%   order of mixin application is relevant in often subtitle and undesired
%   ways. constraints. This hinders their usability and their ability of resolving
%   conflicts: the linearisation (total ordering) of mixin inheritance cannot
%   provide a satisfactory resolution in all cases and restricts the flexibility
%   of mixin composition.  To fight those limitations, an algebra of mixin
%   operators is introduced~\cite{ancona2002calculus}, but this raised the
%   complexity of the approach, especially when constructors and fields are
%   considered~\cite{marco09FJigsaw}.  Our approach does not have the
%   linearisation problem, in that the semantics of Java \textbf{extends} clause
%   is unordered and symmetric.
% \item Simplifying the mixin algebras approach, traits draw a strong line between
%   units of reuse (traits) and object factories (classes) In this model,
%   traits~\cite{scharli03traits} units of reusable code, containing only methods
%   as reusable functionalities. Thus, no state/state initialization is
%   considered.

%   Classes act as object factories, requiring functionalities from multiple
%   traits.  Traits offers a trait algebra with operations like sum, alias and
%   exclusion, provided for explicit conflict resolution.

%   Concluding, (pure) traits do not allow state and they do not offers any reuse
%   instrument to ensure that state is coherently initialized when finally defined
%   in classes.  Traits can't be instantiated. This model requires two concepts
%   (traits and classes) to coexist and cooperate.

%   Some authors see this as good language design fostering good software
%   development by helping programmers to think about the structure of their
%   programs.  However, other authors see the need of two concepts and the absence
%   of state as drawbacks of this model. Our approach takes interfaces as units of
%   reuse, and meanwhile generates factory method for instantiation.

% \item C++ and Scala also try to provide solutions to multiple inheritance, but
%   both suffer from object initialization problems. Virtual inheritance in C++
%   provides another solution to multiple inheritance (especially the diamond
%   problem by keeping only one copy of the base class)~\cite{ellis1990annotated},
%   however suffers from object initialization problem as pointed out by Malayeri
%   et al.~\cite{malayeri2009cz}. It bypasses all constructor calls to virtual
%   superclasses, which would potentially cause serious semantic error. Scala
%   solution (very similar to linearised mixins, but misleadingly called traits in
%   the language) avoids this problem by disallowing constructor parameters,
%   causing no ambiguity in cases such as diamond problem.  This approach has
%   limited expressiveness, and suffers from all the problems of linearised mixin
%   composition.
%   Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
%   \Q@ class A use B,C {...new methods...}@\\*
%   can be expressed in python as \\*
%   \Q@ class Aux: ...new methods...@\\*
%   \Q@ class A(B,C,Aux): pass@
%   Our approach not only does not involve linearised mixin problem, but also
%   support proper constructor mechanism.
% \end{enumerate}

@@

We can model certain types of language extensions with annotations 
only, but those extensions do not introduce new syntax: they 
merely do automatic code generation. 
\end{comment}
\subsection{Meta-programming Competes with Language extensions}
The most obvious solution to features addition to a language is language
extension.  It is often implemented as syntactic extensions that can be
desugared to the base language. For example, the Scala compiler was extended to
directly supports XML syntax. However, this approach does not support combining
multiple extensions into one. We are de facto creating a fork in the language,
and rarely the new fork gain enough traction to become the main language
release.  On this topic we mention SugarJ~\cite{erdweg11sugarj} - a Java-based
extensible language allowing programmers to extend it with custom features by
definitions in meta-DSLs (SDF, Stratego, etc).

On the other side, when the starting language have a flexible enough syntax and
a fast and powerful enough reflection mechanism, we may just need to play with
operator overloading and other language tricks to discover that the language
feature we need can be expressed as a simple library in our language. For
example, consider SQLAlchemy in python.

Java-like languages tend to sit in the middle of two extremes: libraries
can not influence the type system, so many solutions valid in python or other
languages could not be applicable, or may be applicable at the cost of loosing
safety.

Here (compile/load time) code generation comes at the rescue: if for a certain
feature (\mixin in our case) it is possible to use the original language syntax
to \emph{express-describe} any specific instantiation of such feature
(annotating a class and provide getters), then we can insert in the compilation
process a tool that exam and enrich the code before compilation. No need to
modify the original source; for example we can work on temporary files
\yanlin{Here what does temporary files mean? could explain more?}. Java is
a particular good candidate for this kind of manipulation since it already
provide ways to define and integrate such tools in its own compilation process:
in this way there is no need of temporary files, and there is a well defined way
of putting multiple extensions together.

Other languages offer even stronger support to safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time and to generate code/classes that are transparently integrated in the program that is being generated/processed/compiled.
In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or non nonsensical) would never generate ill-typed code. This is roughly equivalent to what we state and manually proof in Lemma 2 for our particular transformation.
Since MetaFjig is not working on annotated classes, there is no direct equivalence on the overall theorem of safety we shown.

\subsection{Formalization of Java8}
We provide a simple and well designed formalization for a subset of Java
including default/static interface methods and object initialization literals
(often called anonymous local inner classes).  A similar formalization was
drafted by Goetz and Field~\cite{goetz12fdefenders} to formalize defender
(default) methods in Java. However this formalization is limited to model
exactly one method inside classes/interfaces.

% As an evidence of the attention and care present in our formalization work,
% while double-checking the behaviour of Java in side cases we have discovered a
% likely bug in the current \texttt{javac}.\marco{refer to before when we explain
%   the issue.}
\yanlin{bug claim removed.}

\section{Future work}\label{sec:futurework}
